Transferability of clinical prediction models for early trauma care

### 2019-08-01 ManuscriptMarkdown.Rmd
Update ManuscriptMarkdown.Rmd

---
title: "Transferability of clinical prediction models for early trauma care"
author: 
- Martin Henriksson
- Dell Saulnier
- Johanna Berg
- Martin Gerdin Wärnberg
date: "July 31, 2019"
documentclass: "article"
output:
  word_document:
    reference_docx: transferstyle.docx
header-includes:
   - \usepackage{hyperref}
bibliography: library.bib
csl: bmcemerg.csl
---
```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)

@@

# Introduction
Trauma, defined as an injury to living tissue by an extrinsic agent [@RN63], 
results in approximatly 458,5 million cases anually across the globe [@RN999], 
and an estimated 3,8 million deaths [@RN1; @RN21]. Each year 9% of the world’s 
deaths are because of trauma, with the leading causes being road traffic accidents, 
suicide and homicide. Predictions indicate the incidence of these causes are 
likely to increase by 2030 [@RN42].

@@

Systems that determine the level of trauma can be based on clinical prediction
models. Models differ in quality and characteristics, but generally perform
well at predicting survival [@RN45]. Many models are developed in a single,
standardized setting, such as a major trauma centre but are then implemented in 
different settings, such as non-trauma centres.

@@

In trauma, misclassification is often referred to as mistriage. Triage refers to
the classification of severity as minor or major trauma. Mistriage can be
subdivided into overtriage, the incorrect classification of a patient with minor
trauma as major trauma, or undertriage, the incorrect classification of a patient
with major trauma as minor trauma. Mistriage can ultimatly lead to decreased patient 
survival, but is also detrimental in terms of patient care and distribution of resources.

Thus, the effect of model transfers between care contexts within a single healthcare system, 
as well as the effect of such transfers on mistriage,have not been studied and represent 
substantial knowledge gaps. Therefore, the aim of this study is to assess how transfers 
of clinical prediction models for early (on arrival to hospital) trauma 
care between different care contexts within a single health system affect mistriage rates.

@@

The trauma care in Sweden is usually clearly defined in major
metropolitan areas: trauma patients are transported directly to a predesignated
hospital with specific trauma competency. In more rural parts of the
country, trauma patients are instead transported to the nearest hospital for
stabilization. Once stabilized, they may then be transported to hospitals with
more advanced trauma care capacity [@RN34]. Most hospitals in Sweden use some
form of system to categorize the level of trauma, usually
based on patient vital signs and injury mechanism, as reported by the EMS, or as
registered in the emergency department [@RN90].

Today 52 of the 55 Swedish hospitals record trauma cases in SweTrau. 
Currently, the registry includes 55 000 cases, and its use is encouraged in 
both academic research and local quality improvement initiatives [@RN28].

## Participants 
The eligibility criteria were patients aged 15 years or above registered in SweTrau 
between 2011 and 2016. The study aimed to assess adult trauma and not
paediatric trauma which differs in physiology, triage and initial care
[@RN46]. The Swedish guidelines for trauma activation define children as age <15
[@RN90].

@@

## Bias
Data analysis was conducted according to a prearranged analysis plan. The
analysis plan and statistical analysis code was finalised using simulated
data. These efforts were taken to avoid confirmation bias. The analysis plan was
reviewed by an experienced statistician and programmer prior to implementation. 
Neither the outcome nor variables were blinded when during analysis, which 
made a structured approach important, to ensure objectivity.

@@

## Statistical methods
### Data sets
The complete SweTrau cohort was split into three overlapping sets of data, each representing a 
plausible care context for model transfer. These data sets were further divided 
into samples, this process is outlined below.

@@

Table 2 shows model validation performance. The model with the lowest validation
mistriage rate was the `r colnames(lowest.val.mistriage)` model with a median
mistriage rate of `r lowest.val.mistriage`. Performance in terms of
undertriage and overtriage rate for the same model was `r val.undertriage`, and
`r val.overtriage` respectively. Worst validation performance (i.e. highest
mistriage rate) was found in the `r colnames(highest.val.mistriage)` sample, with
a model median mistriage rate of `r highest.val.mistriage`.

@@

Due to centralized trauma care, non-metropolitan centres generelly have fewer 
resources, and less experience reciving trauma patients. They may rely 
on clinical prediction models more heavily for accurate trauma triage, and an 
increased mistriage leads to further strain on already limited resources
due to increased overtriage. Increased undertriage could also possibly lead to
increased patient mortality

Recognizing these risks and updating models accordingly should lead
to increased accuracy during trauma patient triage. Potentially saving lives by
minimizing undertriage, and likely save resources by minimizing overtriage. With
registries such as SweTrau, analysis of model transfer could be performed
continuously to optimize models currently in use and advise which model
transfers might be detrimental (or potentially beneficial).

>>>>

---
title: "Transferability of clinical prediction models for early trauma care"
author: 
- Martin Henriksson
- Dell D Saulnier
- Johanna Berg
- Martin Gerdin Wärnberg
date: "July 31, 2019"
documentclass: "article"
output:
  word_document:
    reference_docx: transferstyle.docx
header-includes:
   - \usepackage{hyperref}
bibliography: library.bib
csl: bmcemerg.csl
---
```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)

@@

# Introduction
Trauma, defined as physical injuries to a host by outside objects [@RN63], 
results in approximately 458.5 million cases anually across the globe [@RN999], 
and an estimated 3.8 million deaths [@RN1; @RN21]. Each year, 9% of the world’s 
deaths are the result of trauma, with the leading causes being road traffic accidents, 
suicide, and homicide. Predictions indicate the incidence of these causes is 
likely to increase by 2030 [@RN42].

@@

Systems that determine the level of trauma during early trauma care can be based 
on clinical prediction models. Models differ in quality and characteristics, but 
generally perform well at predicting survival [@RN45]. Many models are developed 
in a single, standardized context, such as a major trauma centre but are then 
implemented in different context, such as non-trauma centres.

@@

In trauma, misclassification is often referred to as mistriage. Triage refers to
the classification of severity as minor or major trauma. Mistriage can be
subdivided into overtriage, the incorrect classification of a patient with minor
trauma as major trauma, or undertriage, the incorrect classification of a patient
with major trauma as minor trauma. Mistriage can ultimatly lead to decreased patient 
survival, and is also detrimental to patient care and distribution of resources [@RN44].

Thus, the effect of model transfers between care contexts within a single healthcare system, 
as well as the effect of such transfers on mistriage, have not been studied and represent 
substantial knowledge gaps. The aim of this study is to assess how transfers 
of clinical prediction models for early trauma care between different care contexts 
within a single health system affect mistriage rates.

@@

Trauma care in Sweden is usually clearly defined in major
metropolitan areas: trauma patients are transported directly to a predesignated
hospital with specific trauma competency. In more rural parts of the
country, trauma patients are instead transported to the nearest hospital for
stabilization. Once stabilized, they may then be transported to hospitals with
more advanced trauma care capacity [@RN34]. Most hospitals in Sweden use some
form of system to categorize the level of trauma, usually
based on patient vital signs and injury mechanism, as reported by the EMS, or as
registered in the emergency department [@RN90].

Today, 52 of the 55 Swedish hospitals record trauma cases in SweTrau. 
Currently, the registry includes 55 000 cases, and its use is encouraged in 
both academic research and local quality improvement initiatives [@RN28].

## Participants 
The eligibility criteria were adult patients aged 15 years or above registered in SweTrau 
between 2011 and 2016. The study aimed to assess adult trauma and not
paediatric trauma which differs in physiology, triage and initial care
[@RN46]. The Swedish guidelines for trauma activation define children as age <15
[@RN90].

@@

## Bias
Data analysis was conducted according to a prearranged analysis plan. The
analysis plan and statistical analysis code was finalised using simulated
data. These efforts were taken to avoid confirmation bias. The analysis plan was
reviewed by an experienced statistician and programmer prior to implementation. 
Neither the outcome nor variables were blinded when during analysis, which 
made a structured approach important to ensure objectivity.

@@

## Statistical methods
### Data sets
The complete SweTrau cohort was split into three overlapping sets of data, each representing a 
plausible care context for model transfer. These data sets were further divided 
into samples:

@@

Table 2 shows model validation performance. The model with the lowest validation
mistriage rate was the `r colnames(lowest.val.mistriage)` model with a median
mistriage rate of `r lowest.val.mistriage`. Performance in terms of
undertriage and overtriage rate for the same model was `r val.undertriage`, and
`r val.overtriage` respectively. The worst validation performance (i.e. highest
mistriage rate) was found in the `r colnames(highest.val.mistriage)` sample, with
a model median mistriage rate of `r highest.val.mistriage`.

@@

Due to centralized trauma care, non-metropolitan centres generally have fewer 
resources and less experience receiving trauma patients. They may rely 
on clinical prediction models more heavily for accurate trauma triage, and 
an increase in overtriage can lead to further strains on already limited 
resources. Increased undertriage could possibly lead to increased patient 
mortality.

Recognizing these risks and updating models accordingly could lead
to increased accuracy during trauma patient triage, potentially saving lives by
minimizing undertriage, and likely save resources by minimizing overtriage. With
registries such as SweTrau, analysis of model transfer could be performed
continuously to optimize models currently in use and advise which model
transfers might be detrimental (or potentially beneficial).
